---
title: "Target Illumination by GWAS Analytics (TIGA)"
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Introduction

Genome wide association studies (GWAS) can reveal important genotype to phenotype
associations, however, data quality and interpretability issues must be addressed.
The TIGA approach enables rational ranking, filtering and interpretation of GWAS
via metrics, methods, and interactive visualization.  Each inferred gene-to-trait
association is evaluated for confidence and relevance, with scores solely derived
from aggregated statistics, linking a protein-coding gene and phenotype. Applicability
and thresholds will depend on use cases.

* Updated to [GWAS Catalog](https://www.ebi.ac.uk/gwas/) format v1.0.2, July 15, 2020.
* Workflow designed for integration with TCRD & Pharos
* RCRAS, Relative Citation Ratio (RCR) Aggregated Score based on iCite-RCR.
* Confidence higher (and provenance/accountability) for `mapped` over `reported` SNP-gene associations.
* BETA not comparable for different units, thus using N_beta counts for TIGA.
* Ensembl IDs (from API) used rather than gene symbols for more accurate gene mapping.
* Prototype RShiny app: <http://unmtid-shinyapps.net/tiga/>

## About NHGRI-EBI GWAS Catalog

GWAS Catalog (<http://www.ebi.ac.uk/gwas/>) studies each have a `study_accession`.
Also are associated with a publication (PubMedID), but not uniquely.
See <https://www.ebi.ac.uk/gwas/docs/fileheaders>.

Key definitions:

	`MAPPED GENE(S)`: Gene(s) mapped to the strongest SNP. If the SNP is located
	within a gene, that gene is listed. If the SNP is intergenic, the upstream
	and downstream genes are listed, separated by a hyphen.

	`REPORTED GENE(S)`*: Gene(s) reported by author
	
	`OR or BETA`: Reported odds ratio or beta-coefficient associated with
	strongest SNP risk allele. Note that if an OR <1 is reported this is
	inverted, along with the reported allele, so that all ORs included in
	the Catalog are >1. Appropriate unit and increase/decrease are included
	for beta coefficients.

	`INITIAL SAMPLE DESCRIPTION`: Sample size and ancestry description for
	stage 1 of GWAS (summing across multiple Stage 1 populations, if applicable)

	`PLATFORM (SNPS PASSING QC)`: Genotyping platform manufacturer used in Stage
	1; also includes notation of pooled DNA study design or imputation of SNPs,
	where applicable.

	`GENOTYPING_TECHNOLOGY`: Genotyping technology/ies used in this study, with
	additional array information (ex. Immunochip or Exome array) in brackets.

*  Buniello A, MacArthur JAL, Cerezo M, Harris LW, Hayhurst J, Malangone C, et al. The NHGRI-EBI GWAS Catalog of published genome-wide association studies, targeted arrays and summary statistics 2019. Nucleic Acids Res. 2019;47: D1005–D1012.


```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r message=FALSE, warning=FALSE}
library(readr, quietly=T)
library(data.table, quietly=T)
library(plotly, quietly=T)
```

# Read files

From GWAS Catalog, TCRD, and EFO.

```{r}
ifile_gwas <- "data/gwascat_gwas.tsv"
ifile_assn <- "data/gwascat_assn.tsv"
ifile_snp2gene <- "data/gwascat_snp2gene.tsv"
ifile_trait <- "data/gwascat_trait.tsv"
ifile_icite <- "data/gwascat_icite.tsv"
ifile_tcrd <- "data/tcrd_targets.tsv"
ifile_efo <- "data/efo.tsv"
ifile_efo2do <- "data/ebi-oxo_efo2doid_mappings_maxd-2.tsv"
ifile_efo_gwas <- "data/efo_sub_gwas.tsv"
gt_statsfile <- "data/gt_stats.tsv.gz"
gwas <- read_delim(ifile_gwas, "\t", col_types=cols(.default=col_character(), 
    DATE=col_date(format="%Y-%m-%d"), DATE_ADDED_TO_CATALOG=col_date(format="%Y-%m-%d"),
    ASSOCIATION_COUNT=col_integer(), study_N=col_integer()))
setDT(gwas)
assn <- read_delim(ifile_assn, "\t", col_types=cols(.default=col_character(), INTERGENIC=col_logical(), `P-VALUE`=col_double(), PVALUE_MLOG=col_double(), OR_or_BETA=col_double(), UPSTREAM_GENE_DISTANCE=col_integer(), DOWNSTREAM_GENE_DISTANCE=col_integer(), DATE=col_date(format="%Y-%m-%d"), DATE_ADDED_TO_CATALOG=col_date(format="%Y-%m-%d")))
setDT(assn)
snp2gene <- read_delim(ifile_snp2gene, "\t", col_types=cols(.default=col_character(),
    REPORTED_OR_MAPPED=col_factor(c("r","m","md","mu"))))
setDT(snp2gene)
trait <- read_delim(ifile_trait, "\t", col_types=cols(.default=col_character()))
setDT(trait)
tcrd <- read_delim(ifile_tcrd, "\t", col_types=cols(.default=col_character()))
setDT(tcrd)
```

# Genome Wide Association Studies (GWAS)

### Counts by year, platform

Grouped by vendor (first in list if multiple), though technologies may have evolved for a given
vendor.

```{r}
sprintf("Studies: %d; traits: %d ; PMIDs: %d", uniqueN(gwas$STUDY_ACCESSION), uniqueN(gwas$DISEASE_TRAIT), uniqueN(gwas$PUBMEDID))
gwas[, `:=`(YEAR = format(DATE,"%Y"), PLATFORM = sub("[, ].*$", "", `PLATFORM_[SNPS_PASSING_QC]`))]
gwas_platforms <- gwas[, .(.N), by=c("YEAR", "PLATFORM")]
setorder(gwas_platforms, YEAR, PLATFORM)
gwas_platforms_cast <- dcast(gwas_platforms, YEAR ~ PLATFORM, value.var="N", fill=0)
plot_ly(gwas_platforms_cast, x=~YEAR, y=~NR, name="Not Reported",
                type="scatter", mode="markers", fill="tonexty") %>%
#  add_trace(y = ~Perlegen, name="Perlegen") %>%
  add_trace(y = ~Affymetrix, name="Affymetrix") %>%
  add_trace(y = ~Illumina, name="Illumina") %>%
  layout(title = 'GWAS counts by year &amp; vendor',
         xaxis = list(title="", showgrid=F, range=c(2005, 2019), dtick=1, tickangle=45),
         yaxis = list(title="", showgrid=F),
         margin = list(t=100, l=100),
         font = list(family="monospace", size=18),
         legend = list(x=.1, y=.7), showlegend=T) %>%
  add_annotations(text=("N = count(studies)"), showarrow=F, x=0.05, y=1, xref="paper", yref="paper")
```

### Sample size

Statistical power and strength of evidence is dependent on size of the cohort.
Sample size box plots by year.

```{r}
plot_ly(gwas, x=~YEAR, y=~study_N, type="box") %>%
    layout(title="GWAS sample sizes",
         xaxis=list(title="", tickangle=45), yaxis=list(title="Study_N", type="log"),
         margin = list(t=100, l=100),
         font = list(family="monospace", size=18)) %>%
  add_annotations(text=sprintf("median:%d  mean:%d  range:[%d-%.2g]", median(gwas$study_N, na.rm=T), round(mean(gwas$study_N, na.rm=T)), min(gwas$study_N, na.rm=T), max(gwas$study_N, na.rm=T)), showarrow=F, x=0.5, y=1, xref="paper", yref="paper")
```


```{r}
bins <- c(0, 5e2, 1e3, 1e4, 5e4, 1e5, Inf)
minN = bins[1:length(bins)-1]
maxN = bins[2:length(bins)]
gwas_counts <- data.table(minN=minN, maxN=maxN, count = NA)
for (i in 1:nrow(gwas_counts)) {
  gwas_counts$count[i] <- uniqueN(gwas[(study_N>=minN[i] & study_N<maxN[i]), STUDY_ACCESSION])
}
gwas_counts[, name := sprintf("(%6g <= N < %6g)", minN, maxN)]
gwas_counts[, pct := 100 * count / sum(count)]
write_delim(gwas_counts, "data/gwas_counts_by_samplesize.tsv", "\t")
knitr::kable(gwas_counts[, .(Sample_size = name, Study_count = count, Pct = pct)], caption="GWAS sample sizes", digits=1)
```

### Journals

```{r}
gwas_journals <- gwas[, .(N_gwas = .N, N_pmid = uniqueN(PUBMEDID), N_assn = sum(ASSOCIATION_COUNT)), by="JOURNAL"]
setorder(gwas_journals, -N_gwas)
write_delim(gwas_journals, "data/gwas_counts_by_journal.tsv", "\t")
knitr::kable(gwas_journals[1:12], caption="Top journals by N_gwas")
```

# Relative Citation Ratio (RCR)

RCR annotations from iCite API; PMIDs from GWASCatalog.
New publications may lack RCR. Should we impute RCR=median as reasonable prior?

```{r warning=FALSE}
icite <- read_delim(ifile_icite, "\t", col_types=cols(.default=col_character(),
                       relative_citation_ratio=col_double(), field_citation_rate=col_double(),
                       citation_count=col_integer(), nih_percentile=col_double(),
                       expected_citations_per_year=col_double(), citations_per_year=col_double(),
                       year=col_integer()))
gwas_icite <- merge(gwas[, .(STUDY_ACCESSION, PUBMEDID)], icite, by.x="PUBMEDID", by.y="pmid", all.x=T, all.y=F)
qtl <- quantile(gwas_icite$relative_citation_ratio, seq(0,1,.1), na.rm=T)
anns <- c(sprintf("N_pmid = %d", nrow(gwas_icite)), sprintf("mean = %.1f ; median = %.1f ; max = %.1f", mean(gwas_icite$relative_citation_ratio, na.rm=T), median(gwas_icite$relative_citation_ratio, na.rm=T), max(gwas_icite$relative_citation_ratio, na.rm=T)), sprintf("90%%ile = %.1f", qtl["90%"]), ("(Plot truncated at 25.)"))
plot_ly(gwas_icite, x=~relative_citation_ratio, type="histogram", marker=list(color="#22AA22")) %>%
  layout(title = 'Relative Citation Ratio (RCR) distribution',
         xaxis = list(title="RCR", type="normal", range=c(0,25), showgrid=T),
         yaxis = list(title="", showgrid=T),
         margin = list(t=100, l=100, r=100),
         font = list(family="monospace", size=18)) %>%
  add_annotations(text=paste0(anns, collapse="<br>"), showarrow=F, x=0.5, y=.9, xref="paper", yref="paper")
```

RCR for top GWAS journals by paper count

```{r warning=FALSE}
gwas_icite[!(journal %in% gwas_icite[, .N, by=journal][order(-N)][1:10, journal]), journal := "Other"]
wordwrap <- function(v, len) sapply(sapply(v, strwrap, len), paste, collapse="<br>")
plot_ly(gwas_icite, 
        x=~year, 
        y=~relative_citation_ratio, 
        color = ~journal, 
        type="box",
        text=sprintf("%s: %s, <i>%s</i> (%d)", gwas_icite$PUBMEDID, wordwrap(gwas_icite$title, 40), gwas_icite$journal, gwas_icite$year)) %>%
  layout(title = 'iCite Relative Citation Ratio<br>Top journals by GWAS Publication count',
         xaxis = list(title="", range=c(2015, 2019), showgrid=F, dtick=1, tickangle=45, tickfont=list(size=14)),
         yaxis = list(title="RCR", range=c(0,25), showgrid=T),
         margin = list(t=100, l=100, r=100),
         font = list(family="monospace", size=18),
         boxmode="group",
         legend=list(x=1, y=.9, font=list(size=12)), showlegend=T)
```

# Associations (SNP to trait)

```{r}
sprintf("Associations total: %d ; SNPs: %d ; traits: %d ; PMIDs: %d", nrow(assn), uniqueN(assn$SNPS), uniqueN(assn$DISEASE_TRAIT), uniqueN(assn$PUBMEDID))
```

## CONTEXT aka functionalClass

```{r}
funClass <- assn[, .N, by="CONTEXT"]
funClass <- unique(funClass[grepl(";", CONTEXT), `:=`(CONTEXT = "Multiple", N = sum(N))])
setorder(funClass, -N)
knitr::kable(funClass[N>=10], caption="SNP functional classes (aka CONTEXT), N>=10")
plot_ly(type="pie", values=funClass$N, labels=paste0(funClass$CONTEXT, " (", funClass$N, ")"), textinfo="label+percent", textfont=list(size=12) ) %>%
  layout(title="SNP functional classes (aka CONTEXT)",
         margin = list(t=100, b=100),
         font=list(family="monospace", size=18), showlegend=F)
```

## (UP|DOWN)STREAM_GENE_DISTANCE

```{r}
sprintf("Associations with UPSTREAM_GENE_DISTANCE: %d (%.1f%%)", assn[!is.na(UPSTREAM_GENE_DISTANCE), .N], 100*assn[!is.na(UPSTREAM_GENE_DISTANCE), .N]/nrow(assn))
sprintf("Associations with DOWNSTREAM_GENE_DISTANCE: %d (%.1f%%)", assn[!is.na(DOWNSTREAM_GENE_DISTANCE), .N], 100*assn[!is.na(DOWNSTREAM_GENE_DISTANCE), .N]/nrow(assn))
qtl <- quantile(assn$UPSTREAM_GENE_DISTANCE, na.rm=T)
sprintf("UPSTREAM: %4sile: %d\n", names(qtl), as.integer(qtl))
qtl <- quantile(assn$DOWNSTREAM_GENE_DISTANCE, na.rm=T)
sprintf("DOWNSTREAM: %4sile: %d\n", names(qtl), as.integer(qtl))
```


```{r}
xmax <- 1e5
xvals <- seq(0, xmax, by=1000)
annos <- c(
	sprintf("N_total = %d", assn[, .N]),
	sprintf("N_[BP_up>%.1g] = %d", xmax, assn[UPSTREAM_GENE_DISTANCE>xmax, .N]),
	sprintf("N_[BP_down>%.1g] = %d", xmax, assn[DOWNSTREAM_GENE_DISTANCE>xmax, .N]))
plots <- list(
  plot_ly(alpha=0.6) %>%
    add_histogram(name="UPSTREAM", x=assn$UPSTREAM_GENE_DISTANCE, histnorm="probability") %>%
    add_histogram(name="DOWNSTREAM", x=assn$DOWNSTREAM_GENE_DISTANCE, histnorm="probability") %>%
  add_annotations(text=paste0(annos, collapse="<br>"), x=1, xanchor="right", y=.5, yanchor="bottom", xref="paper", yref="paper", showarrow=F) %>%
    layout(yaxis=list(title="probability")),
  plot_ly() %>%
    add_trace(name="Exponential", type="scatter", x=xvals, y=exp(-xvals/4e4*log(2)), mode="lines", line=list(width=5, color='#CC2222')) %>%
    add_trace(name="ReLU", type="scatter", x=xvals, y=1-xvals/xmax, mode="lines", line=list(width=3, color='#777777', dash="dot"))  %>%
  # add_trace(name="Sigmoid", type="scatter", x=xvals, y=exp(-(xvals-xmax/2)/1e4)/(1+exp(-(xvals-xmax/2)/1e4)), mode="lines", line=list(width=5)) %>%
    layout(yaxis=list(title="f(x)"))
  )
subplot(plots, nrows=2, shareX=T, titleX=F, titleY=T, margin=.05) %>%
  layout(title="Association/SNP-GENE_DISTANCE<br>(UP|DOWN)STREAM",
         margin = list(l=80, t=100, r=50),
         xaxis=list(title="BPs", range=c(0,xmax)),
         font=list(family="monospace", size=14),
         legend = list(x=.8, y=.2), showlegend=T)
```


# SNP to gene mappings

* Catalog mapping via __Ensembl Post-GWAS Analysis Pipeline (POSTGAP)__:
   * [POSTGAP Documentation](https://uswest.ensembl.org/Help/View?db=core;id=596)
   * [POSTGAP Wiki](https://github.com/Ensembl/postgap/wiki/How-do-I-use-POSTGAP-output%3F)
   * [POSTGAP GitHub](https://github.com/Ensembl/postgap/) (_"This script is designed to automatically finemap and highlight the causal variants behind GWAS results by cross-examining GWAS, population genetic, epigenetic and cis-regulatory datasets."_)
* `GSYMB`, `MAPPED_GENE` fields may include chromosomal locations or be "intergenic".
* Chromosomal locations (LOC*) may be non-coding RNA, etc.
* Mappings _within_ gene stronger associations than _upstream_ or _downstream_.
* We understand _upstream_ associations stronger than _downstream_ at equal distance.

```{r}
sprintf("snp2gene: total associations: %d ; studies: %d ; snps: %d ; genes: %d ; intergenic associations: %d ; chromosomal location associations: %d",
        nrow(snp2gene), uniqueN(snp2gene$STUDY_ACCESSION), uniqueN(snp2gene$SNP), 
        uniqueN(snp2gene[!grepl("^LOC", GSYMB) & GSYMB != "intergenic", .(GSYMB)]),
        nrow(snp2gene[GSYMB=="intergenic"]), nrow(snp2gene[grepl("^LOC", GSYMB)]))
snp2gene[, REPORTED_OR_MAPPED := sub('^r$', 'reported', REPORTED_OR_MAPPED)]
snp2gene[, REPORTED_OR_MAPPED := sub('^m$', 'mapped_within', REPORTED_OR_MAPPED)]
snp2gene[, REPORTED_OR_MAPPED := sub('^mu$', 'mapped_upstream', REPORTED_OR_MAPPED)]
snp2gene[, REPORTED_OR_MAPPED := sub('^md$', 'mapped_downstream', REPORTED_OR_MAPPED)]
snp2gene_counts <- snp2gene[, .(N_snp = .N), by="REPORTED_OR_MAPPED"]
knitr::kable(snp2gene_counts)
```
# GWAS Catalog API, Ensembl IDs, Ensemble API and annotations

The API provides additional data including Ensembl IDs for ___mapped genes___, from which we query the Ensembl API
for additional annotations including gene ___biotype___, thereby filtering for __protein_coding__.

```{r}
ensbInfo <- read_delim("data/gwascat_Snps_EnsemblInfo.tsv.gz", "\t", col_types = cols(.default=col_character(), version=col_integer(), strand=col_integer(), start=col_integer(), end=col_integer()))
setDT(ensbInfo)
biotypes <- ensbInfo[, .(N_gene = .N), by="biotype"][order(-N_gene)]
knitr::kable(biotypes[, Percent := round(100 * N_gene / sum(N_gene), 1)][N_gene >= 100])
```

# Gene counts


```{r}
setnames(trait, c("STUDY_ACCESSION", "TRAIT", "TRAIT_URI", "efoId", "efo_label"))
trait <- trait[!is.na(trait$TRAIT_URI)]
trait$TRAIT <- iconv(trait$TRAIT, from="latin1", to="UTF-8")
trait[, trait_id := sub("^.*/", "", TRAIT_URI)] #redundant?
sprintf("MAPPED_GENE values: %d", uniqueN(assn$MAPPED_GENE))
#
assn_reported <- assn[, .(STUDY_ACCESSION, `REPORTED_GENE(S)`)]
assn_reported <- unique(assn_reported[, list(GENE=unlist(strsplit(`REPORTED_GENE(S)`, ", *"))), by=STUDY_ACCESSION])
sprintf("REPORTED_GENE values: %d", uniqueN(assn_reported$GENE))
#
```

TCRD mapping.

```{r}
gsyms_tcrd <- unique(tcrd$tcrdGeneSymbol) #mapped
ensgs_tcrd <- unique(tcrd$ensemblGeneId) #mapped
sprintf("TCRD targets: %d ; ENSGs: %d; geneSymbols: %d", nrow(tcrd), length(ensgs_tcrd), length(gsyms_tcrd))
gsyms_tiga <- unique(snp2gene$GSYMB)
ensgs_tiga <- unique(ensbInfo$id)
gsyms_common <- intersect(gsyms_tiga, gsyms_tcrd)
ensgs_common <- intersect(ensgs_tiga, ensgs_tcrd)
sprintf("GSYMBs mapped to TCRD: %d", length(gsyms_common))
sprintf("ENSGs mapped to TCRD: %d", length(ensgs_common))
tcrd <- merge(tcrd, data.table(ensg=ensgs_tiga, in_gwascat=rep(T, length(ensgs_tiga))),
	by.x="ensemblGeneId", by.y="ensg", all.x=T, all.y=F)
tcrd$in_gwascat[is.na(tcrd$in_gwascat)] <- F
tcrd$idgList <- as.logical(tcrd$idgList)
tcrd$TDL <- factor(tcrd$TDL, levels=c("NA", "Tdark", "Tbio", "Tchem", "Tclin"), ordered=T)
knitr::kable(tcrd[(in_gwascat), .(N_mapped_genes = .N), by="TDL"][order(-TDL)], caption="GWAS mapped gene counts by TDL")
```

# Gene-SNP-Study-Trait (G2T) associations

g2t should have one row for each gene-snp-study-trait association.

```{r}
g2t <- unique(snp2gene[, .(GSYMB, SNP, STUDY_ACCESSION)])
g2t <- merge(g2t, assn[, .(SNPS, STUDY_ACCESSION, PVALUE_MLOG, OR_or_BETA, oddsratio, beta)], 
	all.x=T, all.y=F, by.x=c("SNP", "STUDY_ACCESSION"), by.y=c("SNPS", "STUDY_ACCESSION"))

g2t <- merge(g2t, trait, all.x=F, all.y=F, by="STUDY_ACCESSION", allow.cartesian=T)
g2t <- g2t[!is.na(GSYMB)]
g2t <- g2t[!is.na(OR_or_BETA)]
g2t <- g2t[!grepl("(^LOC|^intergenic)", GSYMB)]

sprintf("GTs with pvalue_mlog, g2t: %d ; genes: %d ; traits: %d",
	 nrow(g2t[!is.na(g2t$PVALUE_MLOG),]),
	 uniqueN(g2t$GSYMB[!is.na(g2t$PVALUE_MLOG)]),
	 uniqueN(g2t$TRAIT[!is.na(g2t$PVALUE_MLOG)]))
sprintf("GTs with or_or_beta, g2t: %d ; genes: %d ; traits: %d",
	 nrow(g2t[!is.na(g2t$OR_or_BETA),]),
	 uniqueN(g2t$GSYMB[!is.na(g2t$OR_or_BETA)]),
	 uniqueN(g2t$TRAIT[!is.na(g2t$OR_or_BETA)]))
sprintf("GTs with oddsratio, g2t: %d ; genes: %d ; traits: %d",
	 nrow(g2t[!is.na(g2t$oddsratio),]),
	 uniqueN(g2t$GSYMB[!is.na(g2t$oddsratio)]),
	 uniqueN(g2t$TRAIT[!is.na(g2t$oddsratio)]))
sprintf("GTs with beta, g2t: %d ; genes: %d ; traits: %d",
	 nrow(g2t[!is.na(g2t$beta),]),
	 uniqueN(g2t$GSYMB[!is.na(g2t$beta)]),
	 uniqueN(g2t$TRAIT[!is.na(g2t$beta)]))
```

# GWAS traits and EFO

EFO = Experimental Factor Ontology. Includes GO, Orphanet, PO, Mondo and Uberon classes.
TSV from source OWL. 

```{r warning=FALSE}
efo <- read_delim(ifile_efo, "\t", col_types=cols(.default=col_character()))
setDT(efo)
efo[, source := sub("^.*/", "", source)] #Uri2Id
efo[, target := sub("^.*/", "", target)] #Uri2Id
efo_classes <- efo[node_or_edge=="node"][, `:=`(node_or_edge=NULL, source=NULL, target=NULL)]
efo_classes[, in_gwascat := as.logical(id %in% trait$efoId)]
efo_counts <- efo_classes[, Ontology := sub("_.*$", "", id)][, .(N_in_gwas = sum(in_gwascat), N_total = .N), by=Ontology][order(-N_in_gwas)]
sprintf("EFO total classes: %d; in GWAS: %d", uniqueN(efo_classes$id), sum(efo_counts$N_in_gwas))
knitr::kable(efo_counts[N_in_gwas>0], caption="EFO sources (GWAS-relevant)")
```

## GWAS trait-subclass relationships 

```{r}
efo_sub <- efo[node_or_edge=="edge"][, `:=`(node_or_edge=NULL, id=NULL, comment=NULL, uri=NULL)]
sprintf("EFO classes: %d ; total subclass relationships: %d", uniqueN(efo$id), nrow(efo_sub))
efo_sub[, source_in_gwas := (source %in% trait$efoId)]
efo_sub[, target_in_gwas := (target %in% trait$efoId)]
efo_sub <- efo_sub[(source_in_gwas) & (target_in_gwas)]
efo_sub[, `:=`(source_in_gwas=NULL, target_in_gwas=NULL, label=NULL)]
setnames(efo_sub, old=c("source", "target"), new=c("trait_id", "subclass_id"))
trait_counts <- trait[, .(N_study = uniqueN(STUDY_ACCESSION)), by=c("efoId", "TRAIT")]
efo_sub <- merge(efo_sub, trait_counts[, .(efoId, trait_N_study=N_study, trait_name=TRAIT)], by.x="trait_id", by.y="efoId", all.x=T, all.y=F)
efo_sub <- merge(efo_sub, trait_counts[, .(efoId, subclass_N_study=N_study, subclass_name=TRAIT)], by.x="subclass_id", by.y="efoId", all.x=T, all.y=F)
efo_sub <- efo_sub[, .(trait_id, trait_name, subclass_id, subclass_name, trait_N_study, subclass_N_study)]
sprintf("GWAS trait-subclass pairs: %d", nrow(efo_sub))
setorder(efo_sub, -trait_N_study, -subclass_N_study)
knitr::kable(efo_sub[1:24], caption="Top EFO trait-subclass pairs")
```

## GWAS studies related by EFO subclass links.

```{r}
efo_gwas <- read_delim(ifile_efo_gwas, "\t", col_types=cols(.default=col_character()))
setDT(efo_gwas)
efo_gwas[, `:=`(trait_id = sub("^.*/", "", trait_uri), subclass_id = sub("^.*/", "", subclass_uri))] #Uri2Id
sprintf("GWAS studies related by EFO-subclass: %d", 
        uniqueN(union(efo_gwas[study_accession != study_accession_subclass, study_accession], efo_gwas[study_accession != study_accession_subclass, study_accession_subclass])))
```

## EFO to DOID (Disease Ontology ID)

From EBI Ontology Xref Service (OxO).
One-to-many and many-to-one mappings exist.
Keep only closest mappings, maximum distance=2.

```{r}
efo2do <- read_delim(ifile_efo2do, "\t", col_types=cols(.default=col_character()))
setDT(efo2do)
efo2do[, `:=`(curie_id = sub(":", "_", curie_id, fixed=T), mapped_curie = sub(":", "_", mapped_curie, fixed=T), mapping_source_prefix = NULL, mapping_target_prefix = NULL)]
setnames(efo2do, old=c("curie_id", "label",  "mapped_curie", "mapped_label"), new=c("efoId", "efo_name", "do_id", "doid_name"))
efo2do <- efo2do[, .SD[distance == min(distance)], by=c("efoId", "do_id")] #keep only closest mappings
gwas_efo_ids <- unique(efo_gwas$trait_id)
efo2do_gwas <- efo2do[(efoId %in% gwas_efo_ids)]
setorder(efo2do_gwas, efoId, do_id)
sprintf("GWAS EFO_IDs (Total): %d", length(gwas_efo_ids))
sprintf("GWAS EFO_ID to DO_ID mappings (distance<=2): EFO_IDs: %d, DO_IDs: %d", 
        uniqueN(efo2do_gwas[, efoId]), uniqueN(efo2do_gwas[, do_id]))
sprintf("GWAS EFO_ID to DO_ID mappings (efo_name=doid_name): EFO_IDs: %d, DO_IDs: %d", 
        uniqueN(efo2do_gwas[efo_name==doid_name, efoId]), uniqueN(efo2do_gwas[efo_name==doid_name, do_id]))
sprintf("GWAS EFO_ID to DO_ID mappings (distance=1): EFO_IDs: %d, DO_IDs: %d", 
        uniqueN(efo2do_gwas[distance==1, efoId]), uniqueN(efo2do_gwas[distance==1, do_id]))
sprintf("GWAS EFO_ID to DO_ID mappings (distance=2): EFO_IDs: %d, DO_IDs: %d", 
        uniqueN(efo2do_gwas[distance==2, efoId]), uniqueN(efo2do_gwas[distance==2, do_id]))
knitr::kable(efo2do_gwas[sample(1:nrow(efo2do_gwas), 10)], caption="EBI OXO EFO-DOID (sample)")
```

# GENE-TRAIT stats

Read gt_stats.tsv, built by tiga_gt_stats.R for TIGA. Statistics designed to weigh evidence aggregated across
studies, for each gene-trait association.


```{r}
gt <- read_delim(gt_statsfile, "\t", col_types=cols(.default=col_character(), 
n_study=col_integer(), study_N_mean=col_integer(), n_snp=col_integer(), n_traits_g=col_integer(), n_genes_t=col_integer(), pvalue_mlog_median=col_double(), or_median=col_double(),
mu_score=col_double(), nAbove=col_integer(), nBelow=col_integer(), mu_rank=col_integer()))
setDT(gt)
sprintf("nrow(gt) = %d", nrow(gt))
```

### Multivariate non-parametric ranking via &mu; scores.

Since weighting among variables of merit is subjective, ranking by &mu; scores is
used, based on the concept of non-dominated solutions and Pareto -boundary/-front.
See Wittkowski (2008) and R package `muStat` for details.

```{r}
knitr::kable(gt[trait=="pathological myopia"], caption="Example: pathological myopia")
```

### Focus on traits with more data and evidence, likely high scientific interest. But not too polygenic?

Which traits are most studied, based on N_study? But also, which of these may be too polygenic to
be suitably specific for drug discovery?

```{r}
trait_top <- trait[, .(TRAIT, N_study=uniqueN(STUDY_ACCESSION)), by="efoId"]
trait_top <- unique(merge(trait_top, gt[, .(N_genes=uniqueN(ensemblId)), by="efoId"], by="efoId", all.x=T, all.y=F))
knitr::kable(trait_top[order(-N_study), .(TRAIT, N_study, N_genes)][1:18], caption="Most-studied traits")
knitr::kable(trait_top[order(-N_genes), .(TRAIT, N_study, N_genes)][1:18], caption="Most-polygenic traits")
```

# Prototype TIGA web app.

TIGA: Target Illumination GWAS Analytics

## Plot single trait with all associated genes.

* X-axis: Evidence (`n_study`)
* Y-axis: Effect (`median(OR)`)
* Other variables used in &mu; scores:
 * `N_trait` - traits associated with gene
 * `N_snp` - # SNPs implicating gene
 * `OR` - median of trait-SNP ORs
 * `N_beta` - # supporting beta values
 * `pVal` - median of trait-SNP pValues
 * `RCRAS` - Relative Citation Ratio Aggregated Score 
 * `N_gene` - # genes associated with trait
 * `N_sample` - mean of sample sizes


Color unmapped gray.

Plot for a selected trait:

```{r}
#trait_id <- "EFO_0001360" #T2DM
#trait_id <- "EFO_0000249" #Alzheimer
#trait_id <- "EFO_0000289" #bipolar disorder
#trait_id <- "EFO_0000249" #Alzheimers disease
#trait_id <- "EFO_0000305" #breast carcinoma
#trait_id <- "EFO_0000270" #asthma
#trait_id <- "EFO_0000692" #schizophrenia
#trait_id <- "EFO_0005842" #colorectal cancer
#trait_id <- "EFO_0001663" #prostate carcinoma
#trait_id <- "EFO_0000685" #rheumatoid arthritis
#trait_id <- "EFO_0003761" #unipolar depression
trait_id <- "EFO_0002508" #Parkinsons disease
#trait_id <- "EFO_0001645" #coronary heart disease
#trait_id <- "EFO_0003885" #multiple sclerosis
#trait_id <- "EFO_0000384" #Crohns disease
#trait_id <- "EFO_0004574" #total cholesterol measurement

query_uri <- paste0("http://www.ebi.ac.uk/efo/", trait_id)
query_trait <- gt[efoId==trait_id, first(trait)]
sprintf("%s: (%s) %s", trait_id, query_uri, query_trait)
```

### Rank and filter based on meanRank score.

```{r}
t2g <- gt[efoId==trait_id]
t2g <- t2g[!is.na(ensemblId)]
t2g <- t2g[!is.na(or_median)]
NMAX_PLOT <- 100
t2g[, ok2plot := as.logical(traitMeanRank<=NMAX_PLOT)]
setorder(t2g, traitMeanRank)
message(sprintf("meanRankScore_filter result: %d / %d", sum(t2g[, ok2plot]), nrow(t2g)))
```

### Plot

```{r}
t2g[is.na(geneIdgTdl), geneIdgTdl := "NA"]
t2g$geneIdgTdl <- factor(t2g$geneIdgTdl, levels=c("Tclin", "Tchem", "Tbio", "Tdark", "NA"), ordered=T)
t2g[, markersize := 10*n_study]
t2g[, text := paste0("<b>", geneSymbol, "</b> (", ensemblId, ")<br><b>", geneName, "</b><br>", "Family: ", geneFamily, "<br>", "TDL: ", geneIdgTdl, "<br>", "N_traits=", geneNtrait, " ; N_snp=", n_snp, " ; N_study=", n_study, "study_N = ",
"OR = ", or_median, "; ",  study_N_mean, "; ",
" ; pVal=", sprintf("%.2g",10^(-pvalue_mlog_median)), "RCRAS = ", rcras, "; ", 
"meanRankScore = ", geneMeanRankScore, "; ","meanRank = ", geneMeanRank, "; ")]
plot_ly(type='scatter', mode='markers', data=t2g[(ok2plot)], 
  color=~geneIdgTdl, colors=c("blue", "green", "red", "black", "gray"),
	x=~geneMeanRankScore,
	y=~or_median, marker=list(symbol="circle", size=~markersize), text=~text) %>%
  layout(xaxis=list(title="Evidence (meanRankScore)", zeroline=F, showline=F), 
         yaxis=list(title="Effect (OddsRatio)"), 
        title=paste0("TIGA: ", query_trait, "<br>(", trait_id, ")"),
        margin=list(t=100,r=50,b=60,l=60), legend=list(orientation="h", x=1, y=1, xanchor="right"), showlegend=T, font=list(family="monospace", size=18)) %>%
  add_annotations(text=sprintf("N: %d (%d shown)", nrow(t2g), sum(t2g$ok2plot)), showarrow=F, x=0, y=1.0, xref="paper", yref="paper")
```

Top hits:

```{r}
setorder(t2g, geneMeanRank)
knitr::kable(t2g[1:10][, .(GSYMB=geneSymbol, name=geneName, family=geneFamily, geneIdgTdl, n_study, study_N_mean, rcras, or_median, pvalue_mlog_median, geneMeanRankScore, geneMeanRank)], caption=sprintf("TIGA: %s, top hits", query_trait))
```

